Estimating demographic bias on tests of children's early vocabulary

### 2022-01-13 paper/cogsci.Rmd
writing

---
title: "Words aren't created equal: Investigating bias on the CDI"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

author-information: > 
    \author{{\large \bf Morton Ann Gernsbacher (MAG@Macc.Wisc.Edu)} \\ Department of Psychology, 1202 W. Johnson Street \\ Madison, WI 53706 USA
    \AND {\large \bf Sharon J.~Derry (SDJ@Macc.Wisc.Edu)} \\ Department of Educational Psychology, 1025 W. Johnson Street \\ Madison, WI 53706 USA}

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to accurately measure early language. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
    measures of children's aggregate early language skill. However, CDIs are predominantly 
    comprised of hundreds of vocabulary items, some of which may not be heard (and thus learned) 
    equally often by children of varying backgrounds. Here, we use a database of American English 
    CDIs to identify words that show strong bias for particular groups of children, along the axes
    of sex (male vs. female), race/ethnicity (white vs. non-white), and socioeconomic status 
    (high vs. low). For each axis, we identify dozens of strongly biased items, and show that 
    eliminating these items reduces the expected ability difference between groups. For sex, we
    consider how to propose replacement words that may show less bias, on the basis of their 
    relatively equal frequency in adult speech directed to male and female children.

@@

Researchers, clinicians, and parents have long been fascinated with the surprising speed and variability in the growth of young children's vocabulary.
Children's early vocabulary growth is assumed to reflect not only their exposure to child-directed speech, but also the varying difficulty of different types of words, and individual differences in the aptitude of the child -- including potential language deficits.
Research has uncovered both within-child consistency in development, as well as significant influence from external factors.
For example, @bornstein2016stability found stability in core language skills across 10 years of children's development, despite changes in maternal income and education over the study period.
Yet socioeconomic status (SES) has also been found to be predictive of children's early language skill, and of later educational outcomes [for a review, see @schwab2016].
Biological factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994] -- an sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
However, it is also difficult to measure language skill: in any short recording children are unlikely to use all of the words and constructions that they know, and any comprehensive battery of language measures will likely exhaust children's attention span.

@@

While it is quite reasonable to expect that some CDI words are easier than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less difficult) for different groups of children.
The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring particular one group over another (e.g., rural vs. urban).
Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs. female), socioeconomic status (low- vs. high-SES), and ethnicity (white vs. non-white).

We take the approach put forward by @stenhaug2021treading: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.

DIF is, however, a challenging metric for identifying bias for a couple reasons. 
Firstly, DIF is extremely hard to accurately root out. 
The process of finding DIF involves using the test you used to do IRT analysis in order to look for bias within the very same test. 
Secondly, the presence of DIF does not necessarily indicate the presence of bias; it can indicate that one demographic has a higher average ability level than the other. 
These two issues are the basis upon which we get the Fundamental DIF Identification problem.

This problem can be be understood through a concrete example that is well-known from research on early language learning: consider the fact that females show a larger vocabulary than males across early development [see @frank2021 Ch. 6?; OTHER REFS]. 
The question is whether or not females actually have a higher ability level, or if the tests predominantly have words that are easier for females (i.e., biased).
(It could even be that the test is biased toward males but the language ability of females is large enough to overcome that bias.)
To address this question, we need to know the ability levels of girls and boys in order to confirm that if a word is learned earlier by girls it is simply because of an ability difference rather than a bias inherent to the word. The problem arises from the fact that we measure ability level using the very same test that we are trying to check for biased words. 
If the boys and girls were of the same average language ability, this would not be an issue but given the evidence showing that girls have a higher ability level, we need to know the difference in ability so that when we find DIF for a specific word can be sure it is outside of the expecting DIF that results naturally from their difference in ability.

The outline of this paper is as follows.
First, we will introduce the Wordbank data and the IRT model we use to analyze the data.

@@

We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
Full demographic data are not reported in some datasets contributed to Wordbank: sex is available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, `r nrow(d_demo)-sum(is.na(d_demo$sex))`.

We focused on comparing data between demographic groups on the axes of Sex, Ethnicity and Mother's Education, a proxy for socioeconomic status (henceforth, SES). 
Our data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants.
In terms of Ethnicity data was recorded from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` “Other” participants as well as `r sum(is.na(d_demo$ethnicity))` unreported. Given our distribution of ethnicities, for our analysis we split participants into White (`r eth2_tab["White"]`) and Non-White (`r eth2_tab["Nonwhite"]`). 
Finally, we split participants into high and low SES, of which we had `r ses_tab["high"]` high SES participants, corresponding to mothers' education of some college or higher and `r ses_tab["low"]` Low SES participants. 
<!-- same for eth, SES -->

>>>>

---
title: "Estimating demographic bias on tests of children's early vocabulary"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

author-information: > 
    \author{{\large \bf George Kachergis (kachergis@stanford.edu)}
    \AND {\large \bf Nathan Francis (nathan99@stanford.edu)}
    \AND {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA
    }

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to accurately measure early language. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
    measures of children's aggregate early language skill. However, CDIs are predominantly 
    comprised of hundreds of vocabulary items, some of which may not be heard (and thus learned) 
    equally often by children of varying backgrounds. Here, we use a database of American English 
    CDIs to identify words that show strong bias for particular groups of children, on dimensions
    of sex (male vs. female), race/ethnicity (white vs. non-white), and socioeconomic status 
    (high vs. low). For each dimension, we identify dozens of strongly biased items, and show that 
    eliminating these items reduces the expected ability difference between groups. For sex, we
    consider how to propose replacement words that may show less bias, on the basis of their 
    relatively equal frequency in adult speech directed to male and female children.

@@

Researchers, clinicians, and parents have long been fascinated with the surprising speed and variability in the growth of young children's vocabulary.
Children's early vocabulary growth is assumed to reflect not only their exposure to child-directed speech, but also the varying difficulty of different types of words, and individual differences in the aptitude of the child -- including potential language deficits.
Research has uncovered both within-child consistency in development, as well as significant influence from external factors.
For example, @bornstein2016stability found stability in core language skills across 10 years of children's development, despite changes in maternal income and education over the study period.
Yet socioeconomic status (SES) has also been found to be predictive of children's early language skill, and of later educational outcomes [for a review, see @schwab2016].
Demographic factors are also predictive of language skill: first-born children tend to outpace their siblings, and female children tend to have better language skills than their age-matched male counterparts [@fenson1994] -- a sex-based verbal advantage that continues through high school [see @petersen2018gender for a review].
However, it is also difficult to measure language skill: in any short recording children are unlikely to use all of the words and constructions that they know, and any comprehensive battery of language measures will likely exhaust children's attention span.

@@

While it is quite reasonable to expect that some CDI words are easier than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less difficult) for different groups of children.
The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
On any given test, it is clearly undesirable to have more items favoring one group (say, children from rural households) over another (urban children), as the test will overestimate the ability of test-takers in the former (rural) group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair.
A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring particular one group over another (e.g., rural vs. urban).
Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs. female), socioeconomic status (low- vs. high-SES), and ethnicity (white vs. non-white).

However, DIF is fundamentally difficult to identify for multiple reasons [for an overview, see @stenhaug2021treading].
First, most techniques to identify DIF rely on defining a set of "anchor" test items that are assumed to be equally difficult (i.e., unbiased) for both groups of interest.
Identifying anchor items is at best fraught when there may in fact be a difference in ability between groups (e.g., the female language advantage), and is further confounded when the magnitude of this ability difference is unknown.
A reason that DIF is particularly tricky in measuring children's early language ability is that there is a finite universe of early-learned words to choose from--and we may expect many of them to be biased for various environmental reasons (e.g., children in Florida may not use mittens or skis).
Hence, the presence of DIF on a wide variety of items may not indicate the presence of bias; it could indicate that one demographic has a higher average ability level than the other.

<!--
This problem can be be understood through a concrete example that is well-known from research on early language learning: consider the fact that females show a larger vocabulary than males across early development [see @frank2021 Ch. 6?; OTHER REFS]. 
The question is whether or not females actually have a higher ability level, or if the tests predominantly have words that are easier for females (i.e., biased).
(It could even be that the test is biased toward males but the language ability of females is large enough to overcome that bias.)
To address this question, we need to know the ability levels of girls and boys in order to confirm that if a word is learned earlier by girls it is simply because of an ability difference rather than a bias inherent to the word. The problem arises from the fact that we measure ability level using the very same test that we are trying to check for biased words. 
If the boys and girls were of the same average language ability, this would not be an issue but given the evidence showing that girls have a higher ability level, we need to know the difference in ability so that when we find DIF for a specific word can be sure it is outside of the expecting DIF that results naturally from their difference in ability.
-->

The outline of this paper is as follows.
First, we will introduce the Wordbank data and the IRT model we use to analyze the data.
We then fit the IRT model to each group along each dimension, and examine the parameters for evidence of DIF

@@

We analyze parent-reported Wordbank data from `r nrow(d_demo)` American English CDI: Words & Sentences administrations for children `r min(d_demo$age)` to `r max(d_demo$age)` months of age [@frank2017; @frank2021].
Full demographic data are not reported in some datasets contributed to Wordbank: sex was available for `r nrow(d_demo)-sum(is.na(d_demo$sex))` children, race/ethnicity was available for `r nrow(d_demo)-sum(is.na(d_demo$ethnicity))` children, and mother's education (a proxy for socioeconomic status; SES) was available for `r nrow(d_demo)-sum(is.na(d_demo$ses_group))` children.

The analysis of sex-based differences included CDI administrations from `r sex_tab["Female"]` female and `r sex_tab["Male"]` male children.
The analysis of race/ethnicity-based differences included data from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` "Other" children. 
For this analysis we categorized participants' race as White (`r eth2_tab["White"]`) or Non-White (`r eth2_tab["Nonwhite"]`). 
For the SES-based analysis, we categorized the `r ses_tab["high"]` children whose mothers had at least some college or more as high-SES, and those whose mothers had at most high school (`r ses_tab["low"]` children) as low-SES.

### 2022-01-13 paper/cogsci.tex
writing

\title{Words aren't created equal: Investigating bias on the CDI}

\author{{\large \bf Morton Ann Gernsbacher (MAG@Macc.Wisc.Edu)} \\ Department of Psychology, 1202 W. Johnson Street \\ Madison, WI 53706 USA \AND {\large \bf Sharon J.~Derry (SDJ@Macc.Wisc.Edu)} \\ Department of Educational Psychology, 1025 W. Johnson Street \\ Madison, WI 53706 USA}

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{CSLReferences}%
  {}%
  {\par}

@@

\begin{abstract}
Children's early language skill has been linked to later educational
outcomes, making it important to accurately measure early language.
Parent-reported instruments such as the Communicative Development
Inventories (CDIs) have been shown to provide valid, consistent measures
of children's aggregate early language skill. However, CDIs are
predominantly comprised of hundreds of vocabulary items, some of which
may not be heard (and thus learned) equally often by children of varying
backgrounds. Here, we use a database of American English CDIs to
identify words that show strong bias for particular groups of children,
along the axes of sex (male vs.~female), race/ethnicity (white
vs.~non-white), and socioeconomic status (high vs.~low). For each axis,
we identify dozens of strongly biased items, and show that eliminating
these items reduces the expected ability difference between groups. For
sex, we consider how to propose replacement words that may show less
bias, on the basis of their relatively equal frequency in adult speech
directed to male and female children.

@@

Researchers, clinicians, and parents have long been fascinated with the
surprising speed and variability in the growth of young children's
vocabulary. Children's early vocabulary growth is assumed to reflect not
only their exposure to child-directed speech, but also the varying
difficulty of different types of words, and individual differences in
the aptitude of the child -- including potential language deficits.
Research has uncovered both within-child consistency in development, as
well as significant influence from external factors. For example,
Bornstein, Hahn, \& Putnick (2016) found stability in core language
skills across 10 years of children's development, despite changes in
maternal income and education over the study period. Yet socioeconomic
status (SES) has also been found to be predictive of children's early
language skill, and of later educational outcomes (for a review, see
Schwab \& Lew-Williams, 2016). Biological factors are also predictive of
language skill: first-born children tend to outpace their siblings, and
female children tend to have better language skills than their
age-matched male counterparts (Larry Fenson et al., 1994) -- an
sex-based verbal advantage that continues through high school (see
Petersen, 2018 for a review). However, it is also difficult to measure
language skill: in any short recording children are unlikely to use all
of the words and constructions that they know, and any comprehensive
battery of language measures will likely exhaust children's attention
span.

The MacArthur-Bates Communicative Development Inventories {[}CDIs; L.
Fenson et al. (2007){]} are a set of parent-reported measures of
children's productive and receptive language skills, which produce a
low-cost and reliable way to estimate children's early language skills
(Larry Fenson et al., 1994). CDIs have shown good predictive validity
{[}e.g., Larry Fenson et al. (1994); Bornstein et al.~2012, Duff et
al.~2015{]}. Our focus will be on the vocabulary checklist portion of
the CDI Words \& Sentences (CDI:WS) form, comprised of 680 early-learned
words across 22 categories (e.g., animals, vehicles, action words,
pronouns) selected to assess the productive vocabulary of children 16 to
30 months of age. For each item on the CDI:WS, caregivers are asked to
respond whether the target child has been heard to say (i.e.~produce)
the given item.\footnote{The CDI: Words and Gestures form includes a
  subset of the CDI:WS items and targets children 12 to 18 months of
  age, measuring both comprehension and production.} Children's total
vocabulary score on the CDI:WS is tightly correlated with other facets
of early language (e.g., grammatical competence and gesture), suggesting
that the language system is ``tightly woven'' (Frank, Braginsky,
Yurovsky, \& Marchman, 2021). Due to these desirable properties, CDIs
have been adapted to dozens of languages, and a central repository of
CDI data contributed from all over the world has been created
{[}Wordbank; Frank, Braginsky, Yurovsky, \& Marchman (2017); Frank,
Braginsky, Yurovsky, \& Marchman (2021){]}.

Inspired by the utility and widespread use of the CDI, researchers have
recently been using psychometric models on CDI data to construct short,
adaptive tests to reliably assess language ability using only a small
subset of the CDI items Kachergis, Marchman, Dale, Mankewitz, \& Frank
(2021). These psychometric models typically come from the Item-Response
Theory (IRT) framework (Baker, 2001), which assumes that not only
test-takers (here, children) have normally-distributed ability, but that
items (words on the CDI) have normally-distributed difficulty. The very
efficacy of these IRT-based models depends on words varying in
difficulty, and hence being more/less informative of the ability level
of different individuals. For example, asking whether a 22-month-old
produces the word ``ball'' is far less informative of that child's
language ability than asking whether they produce ``table,'' as 96\% of
22-month-olds can produce the former, while only 47\% produce the
latter.

While it is quite reasonable to expect that some CDI words are easier
than others, and even to use these varying difficulties to predict
variation in children's language ability, the use of psychometric models
highlights the possibility that some CDI items may function differently
(i.e., be more/less difficult) for different groups of children. The
idea that some items on a test may show bias, favoring one group over
another, is known as Differential Item Function (DIF; Holland \& Wainer
(1993){]}. A variety of statistical methods for detecting DIF have been
proposed, and investigations have in several instances identified DIF
for many items on tests favoring particular one group over another
(e.g., rural vs.~urban). Our goal here is to test the items on the
American CDI:WS for DIF along three main axes: sex (male vs.~female),
socioeconomic status (low- vs.~high-SES), and ethnicity (white
vs.~non-white).

We take the approach put forward by Stenhaug, Frank, \& Domingue (2021):
GLIMMER plots based on the 1-parameter logistic are sufficient to
identify bias, without potentially obscuring DIF in a more complex
model's additional parameters.

DIF is, however, a challenging metric for identifying bias for a couple
reasons. Firstly, DIF is extremely hard to accurately root out. The
process of finding DIF involves using the test you used to do IRT
analysis in order to look for bias within the very same test. Secondly,
the presence of DIF does not necessarily indicate the presence of bias;
it can indicate that one demographic has a higher average ability level
than the other. These two issues are the basis upon which we get the
Fundamental DIF Identification problem.

This problem can be be understood through a concrete example that is
well-known from research on early language learning: consider the fact
that females show a larger vocabulary than males across early
development {[}see Frank, Braginsky, Yurovsky, \& Marchman (2021) Ch.
6?; OTHER REFS{]}. The question is whether or not females actually have
a higher ability level, or if the tests predominantly have words that
are easier for females (i.e., biased). (It could even be that the test
is biased toward males but the language ability of females is large
enough to overcome that bias.) To address this question, we need to know
the ability levels of girls and boys in order to confirm that if a word
is learned earlier by girls it is simply because of an ability
difference rather than a bias inherent to the word. The problem arises
from the fact that we measure ability level using the very same test
that we are trying to check for biased words. If the boys and girls were
of the same average language ability, this would not be an issue but
given the evidence showing that girls have a higher ability level, we
need to know the difference in ability so that when we find DIF for a
specific word can be sure it is outside of the expecting DIF that
results naturally from their difference in ability.

The outline of this paper is as follows. First, we will introduce the
Wordbank data and the IRT model we use to analyze the data.

@@

We analyze parent-reported Wordbank data from 5520 American English CDI:
Words \& Sentences administrations for children 16 to 30 months of age
(Frank, Braginsky, Yurovsky, \& Marchman, 2017, 2021). Full demographic
data are not reported in some datasets contributed to Wordbank: sex is
available for 4094 children, 4094.

We focused on comparing data between demographic groups on the axes of
Sex, Ethnicity and Mother's Education, a proxy for socioeconomic status
(henceforth, SES). Our data included 1989 female and 2105 male
participants. In terms of Ethnicity data was recorded from 2202 white,
67 Asian, 222 Black, 131 Hispanic, and 93 ``Other'' participants as well
as 2805 unreported. Given our distribution of ethnicities, for our
analysis we split participants into White (2202) and Non-White (513).
Finally, we split participants into high and low SES, of which we had
4973 high SES participants, corresponding to mothers' education of some
college or higher and 547 Low SES participants.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Baker2001}{}%
Baker, F. B. (2001). \emph{The basics of item response theory}. ERIC.

@@

\leavevmode\hypertarget{ref-R-mirt}{}%
Chalmers, R. P. (2012). {mirt}: A multidimensional item response theory
package for the {R} environment. \emph{Journal of Statistical Software},
\emph{48}(6), 1--29.
http://doi.org/\href{https://doi.org/10.18637/jss.v048.i06}{10.18637/jss.v048.i06}

\leavevmode\hypertarget{ref-fenson1994}{}%
Fenson, Larry, Dale, P. S., Reznick, J. S., Bates, E., Thal, D. J.,
Pethick, S. J., \ldots{} Stiles, J. (1994). Variability in early
communicative development. \emph{Monographs of the Society for Research
in Child Development}, i--185.

\leavevmode\hypertarget{ref-Fenson2007}{}%
Fenson, L., Marchman, V. A., Thal, D. J., Dale, P. S., Reznick, J. S.,
\& Bates, E. (2007). \emph{{M}ac{A}rthur-{B}ates {C}ommunicative
{D}evelopment {I}nventories: User's guide and technical manual (2nd
ed.)}. Baltimore, MD: Brookes.

@@

\leavevmode\hypertarget{ref-kachergis2021cat}{}%
Kachergis, G., Marchman, V. A., Dale, P., Mankewitz, J., \& Frank, M. C.
(2021). Online computerized adaptive tests (CAT) of children's
vocabulary development in english and mexican spanish.

\leavevmode\hypertarget{ref-mayor2019}{}%
Mayor, J., \& Mani, N. (2019). A short version of the
{M}ac{A}rthur-{B}ates {C}ommunicative {D}evelopment {I}nventories with
high validity. \emph{Behavior Research Methods}, \emph{51}(5),
2248--2255.

@@

\leavevmode\hypertarget{ref-schwab2016}{}%
Schwab, J. F., \& Lew-Williams, C. (2016). Language learning,
socioeconomic status, and child-directed speech. \emph{WIREs Cognitive
Science}, \emph{7}, 264--275.
http://doi.org/\href{https://doi.org/10.1002/wcs.1393}{10.1002/wcs.1393}

@@

\end{CSLReferences}

>>>>

\title{Estimating demographic bias on tests of children's early vocabulary}

\author{{\large \bf George Kachergis (kachergis@stanford.edu)} \AND {\large \bf Nathan Francis (nathan99@stanford.edu)} \AND {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, Stanford Unviersity \\ Stanford, CA 94305 USA }

@@

\begin{abstract}
Children's early language skill has been linked to later educational
outcomes, making it important to accurately measure early language.
Parent-reported instruments such as the Communicative Development
Inventories (CDIs) have been shown to provide valid, consistent measures
of children's aggregate early language skill. However, CDIs are
predominantly comprised of hundreds of vocabulary items, some of which
may not be heard (and thus learned) equally often by children of varying
backgrounds. Here, we use a database of American English CDIs to
identify words that show strong bias for particular groups of children,
on dimensions of sex (male vs.~female), race/ethnicity (white
vs.~non-white), and socioeconomic status (high vs.~low). For each
dimension, we identify dozens of strongly biased items, and show that
eliminating these items reduces the expected ability difference between
groups. For sex, we consider how to propose replacement words that may
show less bias, on the basis of their relatively equal frequency in
adult speech directed to male and female children.

@@

Researchers, clinicians, and parents have long been fascinated with the
surprising speed and variability in the growth of young children's
vocabulary. Children's early vocabulary growth is assumed to reflect not
only their exposure to child-directed speech, but also the varying
difficulty of different types of words, and individual differences in
the aptitude of the child -- including potential language deficits.
Research has uncovered both within-child consistency in development, as
well as significant influence from external factors. For example,
Bornstein, Hahn, \& Putnick (2016) found stability in core language
skills across 10 years of children's development, despite changes in
maternal income and education over the study period. Yet socioeconomic
status (SES) has also been found to be predictive of children's early
language skill, and of later educational outcomes (for a review, see
Schwab \& Lew-Williams, 2016). Demographic factors are also predictive
of language skill: first-born children tend to outpace their siblings,
and female children tend to have better language skills than their
age-matched male counterparts (Fenson et al., 1994) -- a sex-based
verbal advantage that continues through high school (see Petersen, 2018
for a review). However, it is also difficult to measure language skill:
in any short recording children are unlikely to use all of the words and
constructions that they know, and any comprehensive battery of language
measures will likely exhaust children's attention span.

The MacArthur-Bates Communicative Development Inventories (CDIs; Fenson
et al., 2007) are a set of parent-reported measures of children's
productive and receptive language skills, which produce a low-cost and
reliable way to estimate children's early language skills (Fenson et
al., 1994). CDIs have shown good predictive validity {[}e.g., Fenson et
al. (1994); Bornstein et al.~2012, Duff et al.~2015{]}. Our focus will
be on the vocabulary checklist portion of the CDI Words \& Sentences
(CDI:WS) form, comprised of 680 early-learned words across 22 categories
(e.g., animals, vehicles, action words, pronouns) selected to assess the
productive vocabulary of children 16 to 30 months of age. For each item
on the CDI:WS, caregivers are asked to respond whether the target child
has been heard to say (i.e.~produce) the given item.\footnote{The CDI:
  Words and Gestures form includes a subset of the CDI:WS items and
  targets children 12 to 18 months of age, measuring both comprehension
  and production.} Children's total vocabulary score on the CDI:WS is
tightly correlated with other facets of early language (e.g.,
grammatical competence and gesture), suggesting that the language system
is ``tightly woven'' (Frank, Braginsky, Yurovsky, \& Marchman, 2021).
Due to these desirable properties, CDIs have been adapted to dozens of
languages, and a central repository of CDI data contributed from all
over the world has been created (Wordbank; Frank, Braginsky, Yurovsky,
\& Marchman, 2017; Frank et al., 2021).

Inspired by the utility and widespread use of the CDI, researchers have
recently been using psychometric models on CDI data to construct short,
adaptive tests to reliably assess language ability using only a small
subset of the CDI items (e.g., Mayor \& Mani, 2019, p.
@kachergis2021cat). These psychometric models typically come from the
Item-Response Theory (IRT) framework (Baker, 2001), which assumes that
not only test-takers (here, children) have normally-distributed ability,
but that items (words on the CDI) have normally-distributed difficulty.
The very efficacy of these IRT-based models depends on words varying in
difficulty, and hence being more/less informative of the ability level
of different individuals. For example, asking whether a 22-month-old
produces the word ``ball'' is far less informative of that child's
language ability than asking whether they produce ``table'', as 96\% of
22-month-olds can produce the former, while only 47\% produce the
latter.

While it is quite reasonable to expect that some CDI words are easier
than others, and even to use these varying difficulties to predict
variation in children's language ability, the use of psychometric models
highlights the possibility that some CDI items may function differently
(i.e., be more/less difficult) for different groups of children. The
idea that some items on a test may show bias, favoring one group over
another, is known as Differential Item Function (DIF; Holland \& Wainer
(1993){]}. On any given test, it is clearly undesirable to have more
items favoring one group (say, children from rural households) over
another (urban children), as the test will overestimate the ability of
test-takers in the former (rural) group -- and not because of any
underlying mean difference in ability between the groups, but simply
because the test is unfair. A variety of statistical methods for
detecting DIF have been proposed, and investigations have in several
instances identified DIF for many items on tests favoring particular one
group over another (e.g., rural vs.~urban). Our goal here is to test the
items on the American CDI:WS for DIF along three main axes: sex (male
vs.~female), socioeconomic status (low- vs.~high-SES), and ethnicity
(white vs.~non-white).

However, DIF is fundamentally difficult to identify for multiple reasons
(for an overview, see Stenhaug, Frank, \& Domingue, 2021). First, most
techniques to identify DIF rely on defining a set of ``anchor'' test
items that are assumed to be equally difficult (i.e., unbiased) for both
groups of interest. Identifying anchor items is at best fraught when
there may in fact be a difference in ability between groups (e.g., the
female language advantage), and is further confounded when the magnitude
of this ability difference is unknown. A reason that DIF is particularly
tricky in measuring children's early language ability is that there is a
finite universe of early-learned words to choose from--and we may expect
many of them to be biased for various environmental reasons (e.g.,
children in Florida may not use mittens or skis). Hence, the presence of
DIF on a wide variety of items may not indicate the presence of bias; it
could indicate that one demographic has a higher average ability level
than the other.

The outline of this paper is as follows. First, we will introduce the
Wordbank data and the IRT model we use to analyze the data. We then fit
the IRT model to each group along each dimension, and examine the
parameters for evidence of DIF

@@

We analyze parent-reported Wordbank data from 5520 American English CDI:
Words \& Sentences administrations for children 16 to 30 months of age
(Frank et al., 2017, 2021). Full demographic data are not reported in
some datasets contributed to Wordbank: sex was available for 4094
children, race/ethnicity was available for 2715 children, and mother's
education (a proxy for socioeconomic status; SES) was available for 5520
children.

The analysis of sex-based differences included CDI administrations from
1989 female and 2105 male children. The analysis of race/ethnicity-based
differences included data from 2202 white, 67 Asian, 222 Black, 131
Hispanic, and 93 ``Other'' children. For this analysis we categorized
participants' race as White (2202) or Non-White (513). For the SES-based
analysis, we categorized the 4973 children whose mothers had at least
some college or more as high-SES, and those whose mothers had at most
high school (547 children) as low-SES.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Baker2001}{}%
Baker, F. B. (2001). \emph{The basics of item response theory}. ERIC.

@@

\leavevmode\hypertarget{ref-R-mirt}{}%
Chalmers, R. P. (2012). mirt: A multidimensional item response theory
package for the R environment. \emph{Journal of Statistical Software},
\emph{48}(6), 1--29. \url{http://doi.org/10.18637/jss.v048.i06}

\leavevmode\hypertarget{ref-fenson1994}{}%
Fenson, L., Dale, P. S., Reznick, J. S., Bates, E., Thal, D. J.,
Pethick, S. J., \ldots{} Stiles, J. (1994). Variability in early
communicative development. \emph{Monographs of the Society for Research
in Child Development}, i--185.

\leavevmode\hypertarget{ref-Fenson2007}{}%
Fenson, L., Marchman, V. A., Thal, D. J., Dale, P. S., Reznick, J. S.,
\& Bates, E. (2007). \emph{MacArthur-Bates Communicative Development
Inventories: User's guide and technical manual (2nd ed.)}. Baltimore,
MD: Brookes.

@@

\leavevmode\hypertarget{ref-kachergis2021cat}{}%
Kachergis, G., Marchman, V. A., Dale, P., Mankewitz, J., \& Frank, M. C.
(2021). Online computerized adaptive tests (cat) of children's
vocabulary development in english and mexican spanish.

\leavevmode\hypertarget{ref-mayor2019}{}%
Mayor, J., \& Mani, N. (2019). A short version of the MacArthur-Bates
Communicative Development Inventories with high validity. \emph{Behavior
Research Methods}, \emph{51}(5), 2248--2255.

@@

\leavevmode\hypertarget{ref-schwab2016}{}%
Schwab, J. F., \& Lew-Williams, C. (2016). Language learning,
socioeconomic status, and child-directed speech. \emph{WIREs Cognitive
Science}, \emph{7}, 264--275. \url{http://doi.org/10.1002/wcs.1393}

@@
